### 2020-08-07 IJHCSSubmission/bp_studyMethod.tex
Updates from Overleaf

%!TEX root = bioPrediction_main.tex
\section{Field Study}
We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress over time and the feasibility of predicting stress, focus, and awakeness based on biometric signals.

\subsection{Participants}
We recruited group of 14 professionals via personal contacts from a large power and automation company. The group is diverse in terms of age, work experience, gender, and work responsibilities. All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment. Office workers are a population that generalizes to a variety of contexts, and including part-time laboratory workers guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.

@@

\subsection{Procedure}
We performed a study over the course of eight weeks. This study includes the collection of the participant's biometric data, computer interaction data, and self-reporting surveys. We informed participants about the study purpose and procedure, handed out biometric sensors, and introduced and explained the self-reporting to the participants. For 10 of the participants, we further installed a computer activity tracker on their company-issued laptops (note: four participants from the original sample declined for privacy reasons).

After the initial setup, participants were asked to fill out 3 surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}). At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.

Keeping participants engaged in an eight-week study can be
difficult. In the second month of the study, largely to incent participants to
continue, we offered participants two one-hour Tai Chi classes per
week (each in the middle of the day) and asked them to attend one
class a week. The choice of Tai Chi provided a link to mindfulness,
a topic of growing interest in communities. Consultations we held
with researchers in psychiatry and psychology suggested Tai Chi as
a good link to mindfulness and an intervention that might alleviate
stress. By recording attendance of participants at the Tai Chi
sessions, we are able to analyze the impact of the sessions
on the results. As a brief summary, the impact of the
sessions was minimal; a result, we present an analysis
of the effects in the Discussion section of the paper.

@@

\subsection{Data Collection}

In this section we describe the two datasets that we collected from each study participant.

\subsubsection{Biometric Sensors}
Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants. The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/articles/213613165} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.

@@

\begin{table}[h!]
\begin{center}
\small\addtolength{\tabcolsep}{-1pt}
\begin{tabular}{l l}
\hline
Biometric Measurement & Units of Measure \\ 
\hline
\textbf{Physical Activity} & \cite{fox1999,aldana1996}\\
\hspace{3mm}Intensity of motion & (No unit)  \\
\hspace{3mm}Energy Expenditure & Calories per second (cal/s)  \\
\hspace{3mm}Step counter & Steps  \\
\hline
\textbf{Heart} & \cite{haapalainen2010psycho,healey2005detecting,mulder1992measurement,haag2004emotion} \\
\hspace{3mm}Heart rate & Beats per Minute (bpm) \\
\hspace{3mm}Blood pulse wave & (No Unit)  \\
\hspace{3mm}Heart rate variability (RMSSD) & Milliseconds (ms) \\
\hspace{3mm}Blood oxygenation  & Percent (\%)  \\
\hspace{3mm}Blood perfusion & (No unit)  \\
\hline
\textbf{Skin} & \cite{healey2005detecting,haag2004emotion} \\
\hspace{3mm}Galvanic skin response &  kOhm  \\
\hspace{3mm}Skin temperature &  Degrees Celsius ($^{\circ}$C)  \\
\hline
\textbf{Respiration} & \cite{mulder1992measurement,healey2005detecting,haag2004emotion,masaoka1997}\\
\hspace{3mm}Respiratory rate & Breaths per Minute (bpm)  \\
\hline
%\textbf{Environmental}\\
%\hspace{3mm}Barometric pressure & Milibar (mbar)  \\
%\hline
\end{tabular}
\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data}
\label{signals}
\end{center}
%\vspace*{-4mm}
\end{table}

\subsubsection{Computer Interaction Data}
To gain a better understanding of our participants day-to-day work activities, we used an open source computer interaction monitor~\cite{Meyer17}. The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13). %Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.

@@

>>>>

%!TEX root = bioPrediction_main.tex
\section{Field Study}
We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress\rev{, focus and awakeness} over time and the feasibility of predicting stress, focus, and awakeness \rev{using} biometric signals.

\subsection{Participants}
We recruited a group of 14 professionals via personal contacts from a large power and automation company. The group is diverse in terms of age, work experience, gender, and work responsibilities. All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment. Office workers are a population that generalizes to a variety of contexts, including part-time laboratory workers, and guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.

@@

\subsection{Procedure}
We performed a study over the course of eight weeks. This study includes the collection of the participant's \rev{biometric data and self-reporting surveys during their work}. \rev{We also collected computer interaction data, but given privacy concerns of some participants chose not to use this data (see Section~\ref{sec:computer_interaction_data}}). We informed participants about the study purpose and procedure, handed out biometric sensors, introduced and explained the self-reporting to the participants, \rev{handed out consent form and ensured informed consent for the study.}%\tf{and asked them to sign a consent form}.%\tfC{DAVE: do you still have these?}. \gm{Can we state that it was reviewed by a University ethics board?}

After the initial setup, participants were asked to fill out three surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}) \rev{and wear the biometric sensor during their work hours}. At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.

Keeping participants engaged in an eight-week study can be
difficult. In the second month of the study, largely to incent participants to
continue, we offered participants two one-hour Tai Chi classes per
week (each in the middle of the day) and asked them to attend one
class a week. The choice of Tai Chi provided a link to mindfulness,
a topic of growing interest in communities. Consultations we held
with researchers in psychiatry and psychology suggested Tai Chi as
a good link to mindfulness and an intervention that might alleviate
stress. By recording attendance of participants at the Tai Chi
sessions, we are able to analyze the impact of the sessions
on the results. As a brief summary, the impact of the
sessions was minimal; we present an analysis
of the effects in Section~\ref{sec:Discussion}.

%\tfC{What's missing is when we asked participant to wear it. Chris: do you know what the usual time window of the Everion data was? we could state that here or in section 2.3. this would mitigate the question on shower usage since I think we only asked them to wear it during working hours, but I'm not sure anymore :(.}

@@

\subsection{Data Collection}
\label{sec:DataCol}
We collected two datasets  from each study participant.

\subsubsection{Biometric Sensors}
Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants. The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/articles/213613165} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.
\rev{A comparative study has  shown that the Everion can be used as a valid proxy for HRV metrics for knowledge workers~\cite{barrios2019a}.}

@@

\begin{table}[h!]
\begin{center}
\small\addtolength{\tabcolsep}{10pt}%-1pt}
\begin{tabular}{l l}
\hline
Biometric Measurement & Units of Measure \\ 
\hline
\textbf{Physical Activity} & \cite{fox1999,aldana1996}\\
\hspace{3mm}Intensity of motion & (No unit)  \\
\hspace{3mm}Energy Expenditure & Calories per second (cal/s)  \\
\hspace{3mm}Step counter & Steps  \\
\hline
\textbf{Heart} & \cite{haapalainen2010psycho,healey2005detecting,mulder1992measurement,haag2004emotion} \\
\hspace{3mm}Heart rate & Beats per Minute (bpm) \\
\hspace{3mm}Blood pulse wave & (No Unit)  \\
\hspace{3mm}Heart rate variability: RMSSD  & Milliseconds (ms) \\
%\multicolumn{2}{l}{\hspace{4mm}\textit{(RMSSD: root mean square of successive heartbeat interval differences)}} \\
\hspace{3mm}Blood oxygenation  & Percent (\%)  \\
\hspace{3mm}Blood perfusion & (No unit)  \\
\hline
\textbf{Skin} & \cite{healey2005detecting,haag2004emotion} \\
\hspace{3mm}Galvanic skin response &  kOhm  \\
\hspace{3mm}Skin temperature &  Degrees Celsius ($^{\circ}$C)  \\
\hline
\textbf{Respiration} & \cite{mulder1992measurement,healey2005detecting,haag2004emotion,masaoka1997}\\
\hspace{3mm}Respiratory rate & Breaths per Minute (bpm)  \\
\hline
%\textbf{Environmental}\\
%\hspace{3mm}Barometric pressure & Milibar (mbar)  \\
%\hline
\end{tabular}
\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data. \rev{\textit{(RMSSD denotes the root mean square of successive heartbeat interval differences)}}}
\label{signals}
\end{center}
%\vspace*{-4mm}
\end{table}

\subsubsection{Computer Interaction Data}\label{sec:computer_interaction_data}
To gain a better understanding of our participants day-to-day work activities, we \rev{asked participants to install} an open source computer interaction monitor~\cite{Meyer17}. The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13).
% Gail: I commented out Thomas' comment as I don't think we need to address it
%\tfC{Probably needs to state something about how it also captured non-work related browsing (see section 3.4)}
%Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.

@@

\rev{We did not ask our participants about their focus levels in the end of the day survey as focus
is more of an in-the-moment aspect than stress and awakeness.}

### 2020-08-07 IJHCSSubmission/bp_threats.tex
Updates from Overleaf

@@

We also perform a grid analysis in our machine learning model
to pick the number of features that is more predictive
of the productivity-related indicators, 
therefore removing the biometric
features that may be unrelated to the indicators.

>>>>

\rev{Due to the small amount of data available to us for the purposes of this study, there is a risk
that our models may be overfitting to some degree. Such overfitting indicates that the results presented in our work may be less than optimal, however any overfitting applies strictly to the testing data and given that there is no overlap between the data used for training and testing, we do not believe this invalidates our results.
We leave further optimization of our approach to future researchers.}

@@

\rev{A
future more thorough statistical analysis of the relationships between
the aspects of interest, such as stress, and the physiological
data may provide deeper insights into the data and
how it might be used in prediction.}

### 2020-08-07 IJHCSSubmission/elsarticle-template.tex
Updates from Overleaf

\presetkeys{todonotes}{inline}{}
% for commenting and making it visible in the pdf (easy to hide after)
\newboolean{hidecomments}
\setboolean{hidecomments}{false}
%\nochangebars
\ifthenelse{\boolean{hidecomments}}
{\newcommand{\cb}[2]{}}
{\newcommand{\cb}[2]{
		\fbox{\bfseries\sffamily\scriptsize#1}
		{\sf\small$\blacktriangleright$ %$\RHD$
			{#2} $\blacktriangleleft$}}} %\LHD$}}}%
\newcommand\tf[1]{\cb{TF}{\textcolor{green}{#1}}}
\newcommand\cs[1]{\cb{CS}{\textcolor{blue}{#1}}}
\newcommand\ms[1]{\cb{MS}{\textcolor{red}{#1}}}
\newcommand\tfC[1]{\todo[color=green,inline]{Thomas: #1}}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Observing and Predicting Knowledge Worker Stress in the Wild}

@@

%% Group authors per affiliation:
%\author{Elsevier\fnref{myfootnote}}
%\address{Radarweg 29, Amsterdam}
%\fntext[myfootnote]{Since 1880.}

%% or include affiliations in footnotes:
%\author[mymainaddress,mysecondaryaddress]{Elsevier Inc}
%\ead[url]{www.elsevier.com}

%\author[mysecondaryaddress]{Global Customer Service
%\corref{mycorrespondingauthor}}
%\cortext[mycorrespondingauthor]{Corresponding author}
%\ead{support@elsevier.com}

%\address[mymainaddress]{1600 John F Kennedy Boulevard, Philadelphia}
%\address[mysecondaryaddress]{360 Park Avenue South, New York}

\begin{abstract}
Knowledge workers face many challenges in the workplace: work is 
fragmented, disruptions are constant, tasks are complex, and work hours can be 
long. These challenges can impact knowledge workers' stress, focus and awakeness, and in turn their interaction with the digital environment, the quality of work 
performed and their productivity in general. We report on a field study with 14
knowledge workers over an eight-week period in which we investigated, using
experience sampling, how the workers experience stress over time. During this field study, we also collected biometric data, which we then used to investigate if it is possible to predict such productivity-related indicators as stress, focus and awakeness, in the moment.
We observed and report on various trends in knowledge worker stress levels
over several weeks and show that machine learning models can be built to
predict stress, focus and awakeness from biometric data.
\end{abstract}

\begin{keyword}
Biometrics, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
\end{keyword}

@@

\input{bp_intro_v4}
\input{bp_studyMethod}
%\input{bp_data}
\input{bp_results_1}
%\input{bp_results_2}
\input{bp_discussion}
\input{bp_relatedWork}
\input{bp_threats}
\input{bp_conclusion}

>>>>

\presetkeys{todonotes}{inline}{}
% for commenting and making it visible in the pdf (easy to hide after)
\newboolean{hidecomments}
\setboolean{hidecomments}{false}
%\nochangebars
\ifthenelse{\boolean{hidecomments}}
{\newcommand{\cb}[2]{}}
{\newcommand{\cb}[2]{
		\fbox{\bfseries\sffamily\scriptsize#1}
		{\sf\small$\blacktriangleright$ %$\RHD$
			{#2} $\blacktriangleleft$}}} %\LHD$}}}%
\newcommand\tf[1]{\cb{TF}{\textcolor{blue}{#1}}}
\newcommand\cs[1]{\cb{CS}{\textcolor{cyan}{#1}}}
\newcommand\ms[1]{\cb{MS}{\textcolor{red}{#1}}}
\newcommand\gm[1]{\cb{GCM}{\textcolor{magenta}{#1}}}
\newcommand\tfC[1]{\todo[color=green,inline]{Thomas: #1}}
\newcommand{\rev}[1]{\textcolor{blue}{#1}}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{\rev{Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild}}

@@

\begin{abstract}
Knowledge workers face many challenges in the workplace: work is 
fragmented, disruptions are constant, tasks are complex, and work hours can be 
long. These challenges can impact knowledge workers' stress, focus and awakeness, and in turn their interaction with the digital environment, the quality of work 
performed and their productivity in general. We report on a field study with 14
knowledge workers over an eight-week period in which we investigated, using
experience sampling, how the workers experience stress, \rev{focus and awakeness} over time. During this field study, we also collected biometric data \rev{including heart- and skin-related measures}, which we then used to investigate if it is possible to predict stress, focus and awakeness, in the moment.
We observed and report on various trends in knowledge worker stress levels
over several weeks and show that machine learning models can be built
\rev{from the data of a single minimally invasive sensor}
to
predict stress, focus and awakeness \rev{accurately, with the abstract concept of focus
the hardest to predict.}
\end{abstract}

\begin{keyword}
Biometrics, \tf{Psycho-Physiological Data}, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
\end{keyword}

@@

\input{bp_intro_v4}
\input{bp_relatedWork}
\input{bp_studyMethod}
%\input{bp_data}
\input{bp_results_1}
%\input{bp_results_2}
\input{bp_discussion}
\input{bp_threats}
\input{bp_conclusion}

### 2020-08-07 UbiCompSubmission/bp_intro_v4.tex
Updates from Overleaf

Our work builds upon previous work and extends it by performing an
eight-week study in the workplace with 14 knowledge 
workers. To the best of our knowledge, this is the 
longest in-situ study performed in a real-world 
environment of knowledge workers to analyze an extensive collection of biometric signals with experience samples examining real-time prediction of stress in the workplace~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}.
Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, have been much more limited in duration, or are not focused on a real-world environment0.

We collected biometric data as well as
experience samples on stress, and two related human aspects: focus and
awakeness. We collected participant reports on focus because the
challenges that contribute to stress levels can also make it hard for
knowledge workers to stay focused and the level of focus
can affect
productivity~\cite{mark2014bored}.  We collected participant reports
on awakeness because a lack of awakeness (sleepiness) can result in
undesired consequences on work~\cite{connor02}.  The participants in
our study perform various job functions for a research and
development group within a single large corporation. For this study they wore a novel, state-of-the-art
biometric armband sensor\footnote{Biovotion Everion
  sensor~\cite{everion}} with low invasiveness that captured heart-,
respiration- and skin-related measures. This modality was chosen to
ease longitudinal deployment in the field. Using machine learning, we
created classifiers and analyzed their ability to predict stress, focus, and awakeness levels. Considering all of these three human aspects helps us to consider whether one aspect might be easier to detect than another. If one aspect is easier to detect, it might serve
as a proxy or an indicator of the presence (or absence) of other aspects.

>>>>

Our work builds upon previous work and extends it by performing an
eight-week study in the workplace with 14 knowledge 
workers. To the best of our knowledge, this is the 
longest in-situ study performed in a real-world 
environment of knowledge workers analyzing an extensive collection of biometric signals with experience samples examining real-time prediction of stress in the workplace~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}.
Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, much more limited in duration, or not focused on a real-world environment.

We collected biometric data as well as
experience samples on stress, and two related human aspects: focus and
awakeness. We collected participant reports on focus because the
challenges that contribute to stress levels can also make it hard for
knowledge workers to stay focused and the level of focus
can affect
productivity~\cite{mark2014bored}.  We collected participant reports
on awakeness because a lack of awakeness (sleepiness) can result in
undesired consequences on work~\cite{connor02}.  The participants in
our study perform various job functions for a research and
development group within a single large corporation. For this study, they wore a novel, state-of-the-art
biometric armband sensor\footnote{Biovotion Everion
  sensor~\cite{everion}} with low invasiveness that captured heart-,
respiration- and skin-related measures. This modality was chosen to
ease longitudinal deployment in the field. Using machine learning, we
created classifiers and analyzed their ability to predict stress, focus, and awakeness levels. Considering all of these three human aspects helps us to consider whether one aspect might be easier to detect than another. If one aspect is easier to detect, it might serve
as a proxy or an indicator of the presence (or absence) of other aspects.

### 2020-08-07 UbiCompSubmission/bp_threats.tex
Extra revissions
Updates from Overleaf

\textbf{External Validity:}
The results of our study may 
 not generalize to a broader population of office workers.
To mitigate this risk, we collected participants
from a wide variety of different departments
with different age ranges, genders, work experience, and 
working in different positions.

@@

\textbf{Internal Validity:}
This study tries to find correlations between
biometric features and different productivity-related indicators (stress, focus, and awakeness).
Nonetheless, biometric signals are influenced by far more
variables than the ones this study comprehends.
Therefore, trying to draw a strong causality between the biometric
features and the productivity-related aspects would be inaccurate.
To mitigate this risk,  we collected the data
in a rote environment and in a regular manner 
to minimize the number of 
external causes that may affect each participant's
biometric signals.

It is possible that the amount of data collected
is not sufficienet
to draw valid conclusions. To address this threat, 
we collected a data for an eight week period, which is
400\% longer than previous studies~\cite{zuger18,Muller16}.

\textbf{Construct Validity:}
A threat to the study is that
there are other factors that might either influence the
human aspects of interest or that were considered but
are unrelated biometric signals.
To mitigate this risk, we used an state-of-the-art
device that captures a large number of highly accurate biometric
signals. We collected the most commonly analyzed
biometrics that historically have shown correlation with 
productivity aspects from each user.
In an effort to maintain this research applicable to real-world environments, we picked this already existing device, even when, as a trade-off, we could not capture more descriptive and more intrusive signals such as SDNN, SCL, SCR, eye tracking, or brain activity.

>>>>

\subsection{External Validity:}
The results of our study may 
 not generalize to a broader population of office workers.
To mitigate this risk, we collected participants
from a wide variety of different departments
with different age ranges, genders, work experience, and 
working in different positions.

@@

\subsection{Internal Validity:}
This study tries to find correlations between
biometric features and different productivity-related indicators (stress, focus, and awakeness).
Nonetheless, biometric signals are influenced by far more
variables than the ones this study comprehends.
Therefore, trying to draw a strong causality between the biometric
features and the productivity-related aspects would be inaccurate.
To mitigate this risk,  we collected the data
in a rote environment and in a regular manner 
to minimize the number of 
external causes that may affect each participant's
biometric signals.

It is possible that the amount of data collected
is not sufficienet
to draw valid conclusions. To address this threat, 
we collected a data for an eight week period, which is
400\% longer than the longest previous studies~\cite{zuger18,Muller16}.

\subsection{Construct Validity:}
A threat to the study is that
there are other factors that might either influence the
human aspects of interest or that were considered but
are unrelated biometric signals.
To mitigate this risk, we used an state-of-the-art
device that captures a large number of highly accurate biometric
signals. We collected the most commonly analyzed
biometrics that historically have shown correlation with 
productivity aspects from each user.
In an effort to maintain this research applicable to real-world environments, we picked this already existing device, even when, as a trade-off, we could not capture more descriptive and more intrusive signals such as SDNN, SCL, SCR, eye tracking, or brain activity.
